---
title: "Introduction to fmridesign"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to fmridesign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.retina = 1
)
```

```{r packages}
library(fmridesign)
library(ggplot2)
```

## Overview

`fmridesign` helps you build fMRI design matrices cleanly and consistently. It covers task regressors (events convolved with HRFs) and baseline/nuisance terms (drift, intercepts, motion). You use a simple formula interface to define models, visualize and inspect them, and then feed the matrices into your GLM workflow.

Key ideas: event models (task effects), baseline models (drift and nuisance), clear naming, and plots that make it easy to spot problems.

## The fMRI Design Matrix

In fMRI analysis, the design matrix (X) is fundamental to the general linear model:

```text
Y = Xβ + ε
```

Where:
- **Y**: Observed BOLD signal time series
- **X**: Design matrix containing predictors
- **β**: Parameter estimates (regression coefficients)
- **ε**: Error term (noise)

The design matrix typically combines task-related predictors (expected responses to events) and nuisance predictors (drift, motion, physiology). Both are essential for reliable inference.

## Package Architecture

Design construction is split into two parts:

- Event models (`event_model()`): specify onsets (and optional durations), choose HRF basis per term, and include factors, continuous modulators, and covariates.
- Baseline models (`baseline_model()`): add drift terms (polynomial/spline/constant), block-wise intercepts, and any nuisance regressors you provide.

## Quick Start: Complete Workflow

Here's a minimal example demonstrating the complete workflow from experimental design to design matrix:

```{r complete_example, fig.alt="Predicted BOLD regressors over time for an example event model (two runs)."}
# 1. Define the temporal structure (2 runs, 150 scans each, TR = 2s)
TR <- 2
sframe <- sampling_frame(blocklens = c(150, 150), TR = TR)

# 2. Create experimental events
set.seed(123)
# Simple two-condition design with 20 events per run
conditions <- rep(c("A", "B"), each = 10, times = 2)
onsets <- c(
  sort(runif(20, 0, 150 * TR - 10)),  # Run 1
  sort(runif(20, 0, 150 * TR - 10))   # Run 2
)
blockids <- rep(1:2, each = 20)

# 3. Build the event model
emodel <- event_model(
  onset ~ hrf(Condition, basis = "spmg1"),
  data = data.frame(
    onset = onsets,
    Condition = factor(conditions),
    block = factor(blockids)
  ),
  block = ~ block,
  sampling_frame = sframe
)

# 4. Build the baseline model
bmodel <- baseline_model(
  basis = "poly",
  degree = 3,
  sframe = sframe
)

# 5. Extract design matrices
X_task <- design_matrix(emodel)
X_baseline <- design_matrix(bmodel)

# 6. Combine into full design matrix
X_full <- cbind(X_task, X_baseline)
dim(X_full)

# 7. Visualize the complete design
plot(emodel)
```

## Understanding the Components

### Sampling Frame

The `sampling_frame` object defines the temporal structure of your fMRI data:

```{r sampling_frame}
# Single run: 200 scans, TR = 2 seconds
sframe_single <- sampling_frame(blocklens = 200, TR = 2)
print(sframe_single)

# Multiple runs: 3 runs with different lengths
sframe_multi <- sampling_frame(blocklens = c(150, 200, 150), TR = 2)
print(sframe_multi)
```

### Event Specification

Events are typically specified with the formula interface:

```{r event_specification}
# Formula interface (recommended)
emodel_formula <- event_model(
  onset ~ hrf(condition) + hrf(RT, basis = "gaussian"),
  data = data.frame(
    onset = c(10, 30, 50, 70),
    condition = factor(c("easy", "hard", "easy", "hard")),
    RT = c(0.5, 0.8, 0.4, 0.9),
    block = factor(c(1, 1, 1, 1))
  ),
  block = ~ block,
  sampling_frame = sampling_frame(100, TR = 2)
)
```

## Common Use Cases

### 1. Block Design

```{r block_design, fig.alt="Block design regressors (task and rest blocks) over time."}
# Block design with 20-second blocks
block_onsets <- seq(0, 280, by = 40)
block_conditions <- rep(c("task", "rest"), length.out = length(block_onsets))
block_durations <- rep(20, length(block_onsets))

emodel_block <- event_model(
  onset ~ hrf(condition),
  data = data.frame(
    onset = block_onsets,
    condition = factor(block_conditions),
    block = factor(rep(1, length(block_onsets)))
  ),
  block = ~ block,
  durations = block_durations,
  sampling_frame = sampling_frame(150, TR = 2)
)

plot(emodel_block)
```

### 2. Rapid Event-Related Design

```{r rapid_event}
# Rapid event-related with jittered ISI
set.seed(456)
n_events <- 60
rapid_onsets <- cumsum(runif(n_events, 2, 6))  # ISI between 2-6s
rapid_conditions <- sample(c("face", "house", "object"), n_events, replace = TRUE)

emodel_rapid <- event_model(
  onset ~ hrf(stimulus),
  data = data.frame(
    onset = rapid_onsets,
    stimulus = factor(rapid_conditions),
    block = factor(rep(1, n_events))
  ),
  block = ~ block,
  sampling_frame = sampling_frame(ceiling(max(rapid_onsets)/2) + 20, TR = 2)
)

# Check design efficiency
cor(design_matrix(emodel_rapid))
```

### 3. Parametric Modulation

```{r parametric, fig.alt="Parametric modulator (RT) regressor and conditions plotted over time."}
# Event-related design with RT modulation
set.seed(789)
n_trials <- 30
pm_onsets <- sort(runif(n_trials, 0, 200))
pm_conditions <- rep(c("congruent", "incongruent"), length.out = n_trials)
pm_RT <- rnorm(n_trials, mean = ifelse(pm_conditions == "congruent", 0.5, 0.7), sd = 0.1)

emodel_parametric <- event_model(
  onset ~ hrf(condition) + hrf(RT),
  data = data.frame(
    onset = pm_onsets,
    condition = factor(pm_conditions),
    RT = scale(pm_RT)[,1],  # Center the parametric modulator
    block = factor(rep(1, n_trials))
  ),
  block = ~ block,
  sampling_frame = sampling_frame(120, TR = 2)
)

# Visualize the parametric modulator
plot(emodel_parametric, term_name = "RT")
```

## Integration with fMRI Analysis

The design matrices created by `fmridesign` can be used with various fMRI analysis approaches:

```{r integration, eval=FALSE}
# Example: Export for use with external software
X <- cbind(design_matrix(emodel), design_matrix(bmodel))

# Save as text file for SPM, FSL, or AFNI
write.table(X, "design_matrix.txt", row.names = FALSE, col.names = TRUE)

# For use with R-based analysis
# Assuming Y is your fMRI time series data
# fit <- lm(Y ~ X - 1)  # -1 removes intercept as it's in the design

# Or use with specialized fMRI packages
# library(fmri)
# results <- fmri_glm(Y, X)
```

## Best Practices

Always inspect your design. Look for sensible timing, HRF shapes that match expectations, and acceptable correlations among regressors. Plan for efficiency (jitter and balance help), and match baseline flexibility to run length (short runs need fewer drift terms than long ones). Include relevant nuisance regressors when available (e.g., motion, physiology, artifact/outlier flags).

## Next Steps

For more detailed information, see the following vignettes:

- `vignette("a_03_baseline_model")`: In-depth coverage of baseline and nuisance modeling
- `vignette("a_04_event_models")`: Comprehensive guide to event model specification
- Visit the [fmrihrf package documentation](https://github.com/bbuchsbaum/fmrihrf) for HRF details

## Getting Help

If you encounter issues or have questions:

1. Check the function documentation: `?event_model`, `?baseline_model`
2. Browse all available HRFs: `list_available_hrfs()`
3. Report issues: [GitHub Issues](https://github.com/bbuchsbaum/fmridesign/issues)
