---
title: "Encode/Recall multiPPI with parametric modulators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Encode/Recall multiPPI with parametric modulators}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
if (!requireNamespace("multiPPI", quietly = TRUE)) {
  stop("multiPPI must be installed to build this vignette.")
}
if (!requireNamespace("fmridesign", quietly = TRUE) ||
    !requireNamespace("fmrihrf", quietly = TRUE)) {
  knitr::knit_exit()
}
```

## Paradigm overview

This vignette demonstrates how **multiPPI** enables comprehensive analysis of
dynamic functional connectivity in complex experimental designs. We analyze
a simulated encode/recall memory paradigm that exemplifies the kind of rich,
multi-factorial experiments common in cognitive neuroscience.

### Experimental design and neuroscientific motivation

The paradigm implements a 2 (task: encode vs. recall) × 11 (video identities)
event-related design with 3-second trials. During encoding, participants view
video stimuli and are instructed to remember them. During recall, participants
retrieve memories of previously encoded videos and rate their vividness on a
continuous scale.

This design addresses fundamental questions about episodic memory and neural
reinstatement. **Neural reinstatement** refers to the hypothesis that successful
memory retrieval involves reactivating the same patterns of brain activity that
were present during initial encoding. In connectivity terms, we expect that
brain regions that coupled strongly during encoding of a particular stimulus
will show similar coupling patterns when that stimulus is later recalled.

The vividness rating provides a **parametric modulator** that allows us to test
whether subjective memory strength correlates with the degree of connectivity
reinstatement. This addresses whether more vivid memories show stronger
reactivation of encoding-related connectivity patterns.

### Analytical approach

The multiPPI framework provides several complementary analyses for dissecting
connectivity patterns in this paradigm:

- **mppi_reinstatement()**: Quantifies similarity between encoding and recall connectivity patterns
- **mppi_axes()**: Decomposes connectivity changes into gain and routing components
- **mppi_modes()**: Identifies principal communication patterns across all conditions
- **mppi_sdppi()**: Analyzes state-dependent connectivity at the trial level
- **mppi_precision_gate()**: Examines effective connectivity in the precision domain

These analyses collectively provide a comprehensive view of how functional
connectivity patterns are modulated by memory encoding, retrieval, and
subjective memory strength.

## Mathematical and conceptual foundations

### Matrix-PPI framework

The **multiPPI** approach extends traditional seed-based psychophysiological
interaction (PPI) analysis to a matrix framework that estimates condition-specific
changes in functional connectivity without requiring a priori seed regions.

For each experimental condition $k$, multiPPI estimates an **interaction matrix**
$M_k$ that captures how the condition modulates pairwise connectivity relative
to baseline. These matrices can be viewed in different scaling modes:

- **Raw scale**: $M_k$ represents the change in covariance structure
- **Normalized scale**: $\Delta\Sigma_k = \frac{M_k}{\text{tr}(\Sigma_0)}$ where $\Sigma_0$ is the baseline covariance
- **Amplitude scale**: Matrices are scaled to unit Frobenius norm for pattern comparison

### Domain options: BOLD, neural, and innovations

multiPPI supports three complementary analysis domains:

- **BOLD domain** (default) estimates connectivity changes directly in the
  observed BOLD time series, yielding interaction matrices \(\Sigma_k = \Sigma_0 + M_k\)
  where $\Sigma_0$ is the baseline covariance and $M_k$ captures condition
  modulations.
- **Neural domain** deconvolves the hemodynamic response function (HRF) to
  approximate neural-level interactions, using discrete cosine transform (DCT)
  bases with data-driven regularization to obtain sticks \(U_k = \mathrm{deconv}(Y_k, h)\).
- **Innovations domain** whitenes the residuals and psychological modulators via
  AR models (leveraging `fmriAR`), producing HRF-agnostic interaction estimates
  well-suited for rapid exploratory passes or paradigms with uncertain HRF shape.

The default baseline (BOLD) keeps the workflow simple, while the neural and
innovations options can be activated via `domain = "neural"` or
`domain = "innovations"` in the `mppi()` call.

### Basis projection and dimensionality

High-dimensional fMRI data benefits from projection to a lower-dimensional basis
to improve statistical power and computational efficiency. Common approaches include:

- **PCA basis**: Projects data to the top $r$ principal components of the baseline covariance
- **ICA basis**: Uses independent component decomposition for spatial patterns
- **Identity basis**: Retains original voxel space (for small, well-defined ROIs)

The choice of basis dimension $r$ reflects a bias-variance tradeoff: larger $r$
preserves more spatial information but may include noise; smaller $r$ provides
better statistical power but may miss important patterns.

### Statistical inference considerations

Connectivity patterns are assessed using several complementary approaches:

- **Permutation testing**: Null distributions generated by permuting trial labels
- **Cross-validation**: Model performance assessed using held-out data
- **Effect size estimation**: Standardized metrics for comparing connectivity magnitudes

The multiPPI helper functions provide specialized summaries that address different
aspects of connectivity analysis, from basic pattern identification to advanced
state-dependent and precision-domain analyses.

## Simulating the design and dataset

### Design rationale and ground truth connectivity

We generate two 200-TR runs (`TR = 1 s`) with realistic temporal structure and
statistical properties. Each of the 22 conditions (11 encode + 11 recall) appears
three times per run, providing sufficient power for connectivity estimation while
maintaining manageable computational complexity.

The simulation establishes **distinct ground truth connectivity patterns** for
encoding and recall conditions:

- **Encode trials**: Emphasize one set of latent coupling patterns (`encode_patterns`)
- **Recall trials**: Emphasize different latent coupling patterns (`recall_patterns`)
- **Vividness modulation**: Recall trials are scaled by vividness ratings, creating
  a parametric relationship between subjective memory strength and connectivity magnitude

This design allows us to test whether multiPPI analyses can successfully:
1. Distinguish between encode and recall connectivity patterns
2. Detect parametric modulation by behavioral measures
3. Identify shared vs. distinct communication patterns across conditions

### Statistical considerations and simulation parameters

**Temporal structure**: Events are spaced 6 TRs apart with 3 TR duration, providing
sufficient temporal separation to avoid overlap while maintaining realistic event
density for cognitive experiments.

**Latent dimensionality**: Neural activity is generated in a 10-dimensional latent
space (`r = 10`), which is then projected to 80 voxels. This reflects the assumption
that high-dimensional neural data arise from lower-dimensional coordinated patterns.

**Noise characteristics**:
- Base neural activity has modest variance (`sd = 0.08`) with stronger task-driven
  modulations (`sd = 0.5` for coupling patterns, `sd = 0.9` for trial-level activity)
- Measurement noise (`sd = 0.3`) is added after projection to voxel space
- Signal-to-noise ratio is realistic for fMRI connectivity studies

**Orthonormal projection**: The loading matrix uses QR decomposition to ensure
orthonormal columns, preventing artificial correlations between voxels that would
confound connectivity estimates.

```{r simulate-data}
library(multiPPI)
library(fmridesign)
library(fmridataset)
set.seed(2024)

# Design parameters ---------------------------------------------------------
TR <- 1
run_lengths <- c(200, 200)
runs <- rep(seq_along(run_lengths), times = run_lengths)
Tn <- length(runs)
ids <- sprintf("%02d", 1:11)
conditions <- c(paste0("enc_vid", ids), paste0("rec_vid", ids))

# Build event table (duration = 3 TR) -------------------------------------
make_onsets <- function(n_events, spacing, offset = 6L) {
  offset + spacing * seq_len(n_events)
}
per_run_events <- length(conditions) * 3 / length(run_lengths)  # 33 per run
stopifnot(per_run_events %% 1 == 0)

onsets_run1 <- make_onsets(per_run_events, spacing = 6L)
onsets_run2 <- make_onsets(per_run_events, spacing = 6L)

trial_table <- expand.grid(task = c("enc", "rec"), id = ids, rep = seq_len(3L))
trial_table$cond <- with(trial_table, paste0(task, "_vid", id))
trial_table <- trial_table[sample(nrow(trial_table)), ]
trial_table$run <- rep(c(1L, 2L), each = per_run_events)
trial_table$onset <- ifelse(trial_table$run == 1L,
                            onsets_run1,
                            run_lengths[1] + onsets_run2)
trial_table$duration <- 3L
trial_table <- trial_table[order(trial_table$run, trial_table$onset), ]

sf <- sampling_frame(blocklens = run_lengths, TR = TR)
event_mod <- event_model(onset ~ hrf(cond), data = trial_table,
                         block = ~ run, sampling_frame = sf)

# Trial-wise vividness for recall events ----------------------------------
vividness <- rep(0, Tn)
rec_rows <- which(trial_table$task == "rec")
rec_vivid <- runif(length(rec_rows), min = 0.4, max = 1)
for (jj in seq_along(rec_rows)) {
  onset <- trial_table$onset[rec_rows[jj]]
  dur <- trial_table$duration[rec_rows[jj]]
  idx <- onset + seq_len(dur) - 1L
  idx <- idx[idx <= Tn]
  vividness[idx] <- rec_vivid[jj]
}

# Baseline (drift + nuisance) ---------------------------------------------
baseline_mod <- baseline_model(basis = "poly", degree = 3,
                               sframe = sf)

design_base <- mppi_model(event_mod, baseline_model = baseline_mod,
                          runs = runs)

# Simulate latent neural activity (r = 10) --------------------------------
r <- 10
U <- matrix(rnorm(Tn * r, sd = 0.08), Tn, r)
encode_patterns <- matrix(rnorm(length(ids) * r, sd = 0.5), length(ids), r)
recall_patterns <- matrix(rnorm(length(ids) * r, sd = 0.5), length(ids), r)
colnames(encode_patterns) <- colnames(recall_patterns) <- paste0("mode", seq_len(r))

for (ii in seq_len(nrow(trial_table))) {
  onset <- trial_table$onset[ii]
  dur <- trial_table$duration[ii]
  idx <- onset + seq_len(dur) - 1L
  idx <- idx[idx <= Tn]
  id_idx <- match(trial_table$id[ii], ids)
  base_drive <- rnorm(length(idx), sd = 0.9)
  if (trial_table$task[ii] == "enc") {
    pat <- encode_patterns[id_idx, ]
    U[idx, ] <- U[idx, ] + outer(base_drive, pat)
  } else {
    pat <- recall_patterns[id_idx, ]
    scale_factor <- 1 + vividness[idx]
    U[idx, ] <- U[idx, ] + outer(base_drive * scale_factor, pat)
  }
}

# Project to voxel space (V = 80) -----------------------------------------
V <- 80
loadings <- matrix(rnorm(V * r), V, r)
loadings <- qr.Q(qr(loadings))
Y <- U %*% t(loadings) + matrix(rnorm(Tn * V, sd = 0.3), Tn, V)

fmri_ds <- fmridataset::matrix_dataset(datamat = Y, TR = TR, run_length = run_lengths)
```

## Fitting multiPPI with vividness modulators

To span the full experimental space we combine the condition sticks with a
set of recall-vividness modulators using `mppi_parametric()`. The resulting
columns are appended to the design matrix before calling `mppi()` in the
default BOLD domain with an automatic PCA basis (r = 25).

```{r fit-mppi}
# Condition main effects and vividness modulators -------------------------
mods <- mppi_parametric(design_base, T = Tn,
                        mods = list(vivid = vividness),
                        select = "^cond_cond.rec",
                        center = TRUE)

design_info <- mppi_design(design_base, T = Tn)
X_task <- cbind(design_info$S, do.call(cbind, mods))
colnames(X_task) <- c(colnames(design_info$S), names(mods))

base_mat <- design_info$nuisance
if (!is.null(base_mat) && ncol(base_mat)) {
  keep_cols <- !grepl('^constant_', colnames(base_mat))
  base_mat <- base_mat[, keep_cols, drop = FALSE]
}
if (is.null(base_mat) || ncol(base_mat) == 0) {
  X_design <- cbind(`(Intercept)` = 1, X_task)
  psych_idx <- seq(2, ncol(X_design))
} else {
  X_design <- cbind(`(Intercept)` = 1, base_mat, X_task)
  psych_idx <- seq(ncol(base_mat) + 2, ncol(X_design))
}

custom_design <- structure(list(
  X = X_design,
  psych_idx = psych_idx,
  runs = design_info$runs
), class = "mppi_design")

fit <- mppi(fmri_ds, custom_design,
            basis = list(type = "pca", r = 25),
            domain = "bold",
            scale = "cov")

print(fit)
```

By construction the psychological regressors now include one column per condition
(`cond_cond.enc_vid??`, `cond_cond.rec_vid??`) and one parametric column per recall condition
(`cond_cond.rec_vid??::vivid`). When you need neural-domain estimates set
`domain = "neural"` and provide a `deconv` list; for an HRF-agnostic alternative,
use `domain = "innovations"`. The fit stores the relevant residual series
(`fit$U`), residualised psychological vectors (`fit$pk`), and evidence summaries
(`fit$evidence`), enabling all downstream helpers regardless of domain.

## Comprehensive connectivity analysis with helper functions

The multiPPI helper functions provide a comprehensive toolkit for analyzing
different aspects of dynamic functional connectivity. Each function addresses
specific questions about how brain networks are modulated by experimental
conditions. We demonstrate these analyses with the encode/recall dataset,
providing detailed explanations of each method's purpose, implementation, and
interpretation.

```{r helper-summaries}
valid <- vapply(fit$Delta, function(M) !all(is.na(M)), logical(1))
fit_sum <- fit
fit_sum$Delta <- fit$Delta[valid]
fit_sum$names <- fit$names[valid]
fit_sum$pk <- fit$pk[valid]
fit_sum$lagged <- fit$lagged[valid]
if (!is.null(fit$variance)) fit_sum$variance <- fit$variance[valid]
if (!is.null(fit$partial)) fit_sum$partial <- fit$partial[valid]
fit_sum$denom <- fit$denom[valid]
fit_sum$AIC_total <- fit$AIC_total[valid]
if (!is.null(fit$BIC_total)) fit_sum$BIC_total <- fit$BIC_total[valid]
fit_sum$evidence <- fit$evidence

### Gain and routing decomposition with mppi_axes()

**mppi_axes()** decomposes connectivity changes along two orthogonal dimensions
that capture complementary aspects of network modulation:

**Gain axis**: Measures whether condition-specific connectivity changes amplify
or attenuate the brain's intrinsic connectivity structure. Gain is computed as
the alignment between connectivity changes and the first eigenvector of the
baseline covariance matrix:

$$\text{gain} = \frac{v_1^T S_k v_1}{||S_k||_F}$$

where $v_1$ is the dominant baseline connectivity pattern, $S_k = 0.5(M_k + M_k^T)$
is the symmetrized interaction matrix, and the denominator normalizes by the
matrix magnitude.

**Routing axis**: Measures directional information flow by comparing forward
and backward temporal interactions. Using lagged interaction matrices $M_k(\tau)$,
routing is computed as:

$$\text{routing} = \frac{E_+ - E_-}{E_+ + E_- + \epsilon}$$

where $E_+ = \sum_{\tau>0} ||M_k(\tau)||_F$ (forward energy) and $E_- = \sum_{\tau<0} ||M_k(\tau)||_F$
(backward energy).

**Interpretation**: Positive gain indicates amplification of intrinsic patterns;
negative gain indicates attenuation. Positive routing suggests feedforward
processing; negative routing suggests feedback processing.

# Gain vs routing summaries (condition rows only) -------------------------
axes <- mppi_axes(fit_sum, lags = -2:2)
head(subset(axes, grepl("^cond_cond.enc_vid", condition)))

### Communication mode extraction with mppi_modes()

**mppi_modes()** identifies the principal axes of communication across all
experimental conditions by performing eigendecomposition of the aggregate
interaction structure.

**Algorithm**: The function computes an aggregate matrix from all condition-specific
interaction matrices:

$$A = \sum_k M_k M_k^T$$

After symmetrization ($A = 0.5(A + A^T)$) and optional ridge regularization,
eigendecomposition yields:

$$A = W\Lambda W^T$$

The top $r$ eigenvectors in $W$ represent the dominant **communication modes**—
spatial patterns that explain the most variance across all connectivity changes.

**Output**: Each condition's connectivity pattern is projected into this mode space:

$$P_k = W^T M_k W$$

This provides a low-dimensional representation that highlights how each condition
utilizes the fundamental communication patterns.

**Interpretation**: Mode eigenvalues indicate the variance explained by each
spatial pattern. Conditions with similar projections $P_k$ engage similar
communication patterns, while dissimilar projections indicate distinct
network configurations.

# Communication modes spanning the normalized interaction energy ---------
modes <- mppi_modes(fit_sum, r = 6)
colnames(modes$vectors)

### Neural reinstatement analysis with mppi_reinstatement()

**mppi_reinstatement()** tests the **neural reinstatement hypothesis**—that
successful memory retrieval involves reactivating connectivity patterns that
were present during initial encoding.

**Algorithm**: For each stimulus identity, the function compares connectivity
patterns between encoding and recall:

1. **Extract patterns**: Identifies interaction matrices $M_{enc,i}$ and $M_{rec,i}$
   for encoding and recall of identity $i$
2. **Vectorize connectivity**: Converts lower triangular elements to vectors:
   $\vec{m}_{enc,i}$ and $\vec{m}_{rec,i}$
3. **Compute similarities**:
   - **Within-identity**: $r_{within,i} = \text{corr}(\vec{m}_{enc,i}, \vec{m}_{rec,i})$
   - **Between-identity**: $r_{between,ij} = \text{corr}(\vec{m}_{enc,i}, \vec{m}_{rec,j})$ for $i \neq j$
4. **Reinstatement statistic**: $R = \text{mean}(r_{within}) - \text{mean}(r_{between})$

**Interpretation**: Values near 1 indicate perfect reinstatement (encoding and
recall patterns are identical within identities but unrelated between identities).
Values near 0 indicate chance-level reinstatement. Negative values suggest
systematic anti-reinstatement.

**Biological significance**: High reinstatement indicates that brain networks
reconfigure similarly during encoding and recall of the same content, supporting
theories of memory as reactivation of encoding-related neural states.

# Encode vs recall reinstatement (exclude parametric terms via pattern) ---
reinstate <- mppi_reinstatement(fit_sum,
                                enc_pattern = "^cond_cond.enc_vid",
                                rec_pattern = "^cond_cond.rec_vid[0-9]{2}$")
reinstate$stat

### State-dependent connectivity with mppi_sdppi()

**mppi_sdppi()** analyzes **state-dependent psychophysiological interactions**—how
the brain's ongoing state immediately before an event influences connectivity
changes during that event.

**Conceptual framework**: Rather than treating each trial independently, this
analysis recognizes that neural responses to stimuli are shaped by the brain's
instantaneous configuration at stimulus onset. Different pre-event states may
lead to different connectivity patterns for the same experimental condition.

**Algorithm**: For each trial, the function:

1. **Pre-event baseline**: Extracts neural activity in the window `[onset - pre_window, onset - 1]`
2. **State characterization**: Computes features of the pre-event covariance structure:
   - **Effective rank**: Dimensionality of ongoing activity
   - **V1 dominance**: Alignment with the dominant connectivity pattern
   - **Sign balance**: Proportion of positive vs. negative connections
3. **Post-event connectivity**: Measures connectivity features during the trial:
   - **Magnitude**: Overall connectivity strength
   - **Gain/routing**: Alignment with intrinsic patterns and directional flow
4. **Predictive analysis**: Models how pre-event state features predict post-event connectivity

**Output**: Trial-level features that can be used for:
- **State-connectivity relationships**: How brain state influences network dynamics
- **Individual differences**: Subject-specific patterns of state-dependent connectivity
- **Behavioral correlations**: Whether state-dependent connectivity relates to performance

**Interpretation**: High predictive accuracy suggests that connectivity patterns
are strongly influenced by ongoing brain dynamics, while low accuracy indicates
more stimulus-driven, state-independent connectivity changes.

# Trial-level state-dependent PPI metrics (recall conditions) -------------
sd_recall <- mppi_sdppi(fit, design_base, select = "^cond_cond.rec_vid",
                        pre_window = 8L, lags = -2:2)
head(sd_recall)

### Precision domain analysis with mppi_precision_gate()

**mppi_precision_gate()** analyzes connectivity changes in the **precision domain**,
providing insights into effective connectivity—the direct connections between
brain regions after removing indirect effects.

**Theoretical foundation**: While covariance-based connectivity (the default
multiPPI analysis) captures all pairwise relationships, precision-based connectivity
captures direct relationships. The precision matrix $\Theta$ is the inverse of
the covariance matrix: $\Theta = \Sigma^{-1}$.

**Algorithm**: For each condition $k$, the function computes precision perturbations:

$$\Delta\Theta_k = -\Theta_0 M_k \Theta_0$$

where $\Theta_0 = (\Sigma_0 + \lambda I)^{-1}$ is the baseline precision matrix
and $\lambda$ is a ridge regularization parameter for numerical stability.

**Key metrics**:
1. **Total precision energy**: $||\Delta\Theta_k||_F$ measures the overall magnitude
   of direct connectivity changes
2. **Off-diagonal ratio**: $\frac{||\Delta\Theta_k^{off}||_1}{||\text{diag}(\Delta\Theta_k)||_1}$
   compares between-region coupling changes to within-region variance changes

**Interpretation**:
- **High total energy**: Condition strongly modulates direct connections
- **High off-diagonal ratio**: Changes primarily affect between-region coupling
  rather than regional variance
- **Low off-diagonal ratio**: Changes primarily affect within-region properties

**Biological significance**: Precision gating reveals how experimental conditions
selectively modulate specific connections in the network while leaving others
unchanged, providing insights into the targeted nature of cognitive control and
attention mechanisms.

# Precision gating summaries ---------------------------------------------
prec_gate <- mppi_precision_gate(fit_sum)
head(prec_gate$summary)

### Model evidence with mppi_evidence()

To quantify how much the PPI regressors improve model fit relative to the
task-only baseline, call `mppi_evidence()`. Positive ΔAIC/ΔBIC values indicate
additional explanatory power from the interaction terms.

```{r evidence}
ev <- mppi_evidence(fit)
head(ev$per_column[, c("component", "delta_AIC", "delta_BIC")])
ev$aggregate
```

## Results interpretation and statistical considerations

The multiPPI analyses provide complementary views of connectivity patterns that
should be interpreted in conjunction with each other and with knowledge of the
experimental design.

### Expected results for the encode/recall paradigm

Given the simulated ground truth, we expect the following patterns:

**Gain and routing (mppi_axes)**:
- **Encode conditions**: May show positive gain (amplification of intrinsic patterns)
  as encoding processes engage widespread network coupling
- **Recall conditions**: Routing patterns may differ from encoding due to different
  information flow requirements (retrieval vs. consolidation)
- **Vividness effects**: Higher vividness recall trials should show stronger
  connectivity magnitudes

**Communication modes (mppi_modes)**:
- Mode eigenvalues should show a few dominant patterns explaining most variance
- Encode and recall conditions should project differently in mode space,
  reflecting distinct communication strategies
- Within-task conditions (different video identities) should cluster together

**Neural reinstatement (mppi_reinstatement)**:
- Positive reinstatement statistic indicates successful pattern reactivation
- Values > 0.3 typically indicate meaningful reinstatement in real data
- The pattern should be stronger for more vivid recalls if the simulation is realistic

**State-dependent connectivity (mppi_sdppi)**:
- Trial-level variability in connectivity features provides insights into
  individual trial dynamics
- Correlations with behavioral measures (vividness) can validate the analysis
- High effective rank indicates complex, distributed pre-event states

**Precision gating (mppi_precision_gate)**:
- Off-diagonal ratios > 1 indicate that connectivity changes dominate variance changes
- Precision energy should be higher for conditions with stronger ground-truth patterns

### Statistical significance and effect sizes

**Multiple comparisons**: With 22 conditions plus parametric modulators, consider
correction for multiple testing when making statistical inferences.

**Effect size interpretation**:
- **Reinstatement**: Values > 0.2 indicate small effects; > 0.5 indicate large effects
- **Gain/routing**: Values should be interpreted relative to the range [-1, 1]
- **Mode projections**: Compare magnitudes across conditions rather than absolute values

**Cross-validation**: For real data, use permutation testing or cross-validation
to establish statistical significance and avoid overfitting.

## Integration with broader analysis pipelines

The multiPPI framework provides a comprehensive foundation for connectivity analysis
that integrates naturally with other neuroimaging approaches.

### Group-level analysis strategies

**Subject-level summaries**: Each analysis produces subject-specific metrics that
can be aggregated across participants:

```r
# Example group-level analysis workflow (not executed here)
group_reinstatement <- data.frame(
  subject = rep(1:N_subjects, each = 1),
  reinstatement = sapply(subject_fits, function(fit) {
    mppi_reinstatement(fit, enc_pattern = "...", rec_pattern = "...")$stat
  })
)

# Group-level statistical test
t.test(group_reinstatement$reinstatement, mu = 0)
```

**Hierarchical modeling**: Trial-level features from `mppi_sdppi()` can be analyzed
using mixed-effects models to account for within-subject dependencies:

```r
# Mixed-effects analysis of state-dependent connectivity
library(lme4)
model <- lmer(magnitude ~ vividness + (1|subject), data = trial_features)
```

### Integration with other methods

**Complementary to activation analysis**: multiPPI connectivity measures complement
traditional GLM activation analysis by revealing how regions coordinate rather
than just which regions activate.

**Dynamic connectivity**: multiPPI can be combined with sliding-window or other
dynamic connectivity approaches to examine temporal evolution of network states.

**Machine learning**: Connectivity features can serve as inputs to classification
or regression models for predicting behavior or clinical outcomes.

### Validation and quality control

**Simulation validation**: The ground-truth simulation approach demonstrated here
provides a template for validating analysis pipelines with known connectivity patterns.

**Cross-modal validation**: Connectivity patterns should relate to anatomical
connectivity (diffusion MRI) and to behavioral measures in theoretically
meaningful ways.

**Reproducibility**: Use consistent seed values, cross-validation schemes, and
standardized preprocessing to ensure reproducible results across studies.

### Computational considerations

**Memory management**: For large datasets, consider using the chunked processing
options in multiPPI to manage memory usage.

**Parallel processing**: Most multiPPI analyses can benefit from parallel computing
for multiple subjects or conditions.

**Basis selection**: The choice of basis dimension involves a bias-variance tradeoff
that should be validated using held-out data or cross-validation.
