---
title: "Resting-State Trait Analyses with multiPPI"
author: "multiPPI team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Resting-State Trait Analyses with multiPPI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
suppressPackageStartupMessages({
  library(multiPPI)
  library(ggplot2)
  library(dplyr)
})
```

## Overview

This vignette demonstrates how to analyze resting-state data for trait-level
inference using **multiPPI**. We will:

1. Simulate a cohort of resting datasets with known trait effects using the
   `mppi_sim_rest_*` helpers.
2. Fit `mppi_rest()` for each subject with endogenous modulators.
3. Extract Gain/Routing/mode features via `mppi_rest_features()` and link them to
   the trait with a simple cross-validated model.
4. Visualize trait predictions and inspect state-specific coupling patterns.

All analyses run entirely in the low-rank basis returned by `mppi_rest()`, so
scaling and reproducibility are controlled by default.

## 1. Simulate Resting Cohort with Trait Effects

We generate ten subjects with a latent trait that modulates the coupling strength
of one endogenous context (`ctx1`). The helper returns both the simulated data
and the ground truth subject-level amplitudes.

```{r simulate-cohort}
set.seed(1123)
cohort <- mppi_sim_rest_cohort(n_subj = 10,
                               contexts = c("ctx1", "ctx2"),
                               amp_base = c(0.30, 0.25),
                               trait_effect = c(0.6, 0.0),
                               amp_noise = 0.02,
                               obs_noise = 0.01,
                               T = 800,
                               V = 60,
                               r = 6)

str(cohort$trait)
cohort$amplitudes[1:5, ]
```

The `trait` vector is scaled (mean zero, unit variance) and linearly modulates
the amplitude of `ctx1`. Context `ctx2` acts as a control without trait
influence.

## 2. Fit Resting multiPPI per Subject

We now run `mppi_rest()` on each simulated subject. The datasets returned by the
simulator already include precomputed low-rank loadings (`loadings`) and the
modulators used to generate the coupling (`modulators`). Passing them directly
into the wrapper avoids recomputation.

```{r fit-rest}
fits <- lapply(cohort$datasets, function(ds) {
  mppi_rest(ds$Y,
            modulators = ds$modulators,
            basis = ds$loadings,
            prewhiten = FALSE,
            domain = "bold")
})

fits[[1]]
```

Each `mppi_fit` stores the interaction matrices (`Delta`) for every context along
with metadata (domain, basis rank, lagged variants). We can inspect an example:

```{r inspect-delta}
delta_1 <- fits[[1]]$Delta[["ctx1"]]
round(delta_1[1:5, 1:5], 3)
```

## 3. Derive Trait Features and Model the Trait

We summarize each fit using `mppi_rest_features()`, which returns a tidy table of
Gain (`G`), Routing (`R`), and mode magnitude (`mag`) per context and subject.

For the simple trait model we use the mean features per subject and run
`mppi_rest_trait_cv()`, a light ridge regression with 5-fold cross-validation.

```{r trait-features}
feat_long <- mppi_rest_features(fits, aggregate = "none")
head(feat_long)

feat_mean <- mppi_rest_features(fits, aggregate = "mean")
feat_mean

cv_out <- mppi_rest_trait_cv(feat_long, y = setNames(cohort$trait, seq_along(fits)), k = 5)
cv_out$cv_r2
```

Cross-validated $R^2$ quantifies how much variance in the trait is explained by
our normalized coupling features. We can visualize predicted vs observed trait
values:

```{r trait-plot, fig.height=4, fig.alt='Scatterplot comparing observed and cross-validated predicted trait values'}
pred_df <- data.frame(
  subj = as.integer(names(cv_out$pred)),
  observed = cv_out$observed,
  predicted = cv_out$pred
)

ggplot(pred_df, aes(observed, predicted)) +
  geom_point(size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  labs(title = "Cross-validated trait prediction",
       x = "Observed trait",
       y = "Predicted trait") +
  theme_minimal()
```

## 4. Inspect Context-Specific Coupling Patterns

Because we know the ground-truth coupling templates from the simulator, we can
compare recovered interaction matrices against the truth. `mppi_sim_rest_dataset`
stashes the normalized templates under `truth$coupling`.

```{r compare-coupling}
truth_ctx1 <- cohort$datasets[[1]]$truth$coupling[["ctx1"]]
recovered_ctx1 <- fits[[1]]$Delta[["ctx1"]]

cor(as.vector(truth_ctx1), as.vector(recovered_ctx1))
```

The high correlation indicates that multiPPI recovers the context-conditioned
interaction pattern in the low-rank space. We can also visualize the heatmap for
one subject:

```{r heatmap, fig.height=4, fig.alt='Heatmap of recovered ctx1 interaction matrix for a representative subject'}
mat_df <- as.data.frame(as.table(recovered_ctx1))
colnames(mat_df) <- c("row", "col", "value")

ggplot(mat_df, aes(row, col, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C") +
  coord_equal() +
  labs(title = "Recovered coupling matrix (ctx1)", x = "Component", y = "Component") +
  theme_minimal()
```

## 5. Extending to Real Data

For real resting-state datasets, the workflow follows the same structure:

1. Construct or load an `fmri_dataset` and select a basis (`basis = list(type = "pca", r = 100)` is a good default).
2. Let `mppi_rest()` build endogenous modulators using families such as
   `states`, `bursts`, and `envelopes` (set `include = ...`).
3. Extract subject-level features via `mppi_rest_features()` and model the trait
   with `mppi_rest_trait_cv()` or a custom pipeline.
4. Optionally back-project `Delta` matrices to ROI/voxel maps using existing
   helpers to interpret spatial patterns.

Because multiPPI forms interactions in the low-rank neural or innovations domain
and normalizes outputs, the features are robust to between-subject scaling
differences—an essential property for trait studies.

## Session Info

```{r session-info}
sessionInfo()
```
